---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# categorizeQuantiles

<!-- badges: start -->
<!-- badges: end -->

## Description

The goal of categorizeQuantiles is to provide tools for categorizing numeric variables in data frames or tibbles into quantile-based categories. This is especially useful for grouping continuous data into discrete categories based on quantiles, making it easier to analyze and visualize patterns within categorical groups.

## Installation

You can install the development version of categorizeQuantiles from [GitHub](https://github.com/) with:

```r
# in R, make sure that the devtools package is installed then run 
# install.packages("devtools")
devtools::install_github("stat545ubc-2024/categorizeQuantiles")
```


## Example

In this part, let's use a simple tibble and the *penguins* dataset from the *palemerpenguins* package to show the correct usages.

Firstly, load the package and other library:
```{r}
library(categorizeQuantiles)
```

#### 1. Using a Simple Tibble
This example divides the **score** column into two categories: "low", "high".
```{r}
# Define a simple tibble
simple_data <- tibble::tibble(
    id = c("1", "2", "3", "4", "5", "6", "7", "8", "9","10"),
    score = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90)
)

# Display the tibble
print(simple_data)

# Apply categorize_by_quantiles to categorize "score" into two categories
data_categorized_1 <- categorize_by_quantiles(
    numerical_data = simple_data,
    numerical_var_name = "score",
    categorical_labels = c("low", "high"),
    num_categories = 2
)
# Display the result
print(data_categorized_1)
```
#### 2. Using a simple data frame
This example divides the **score** column into four categories: "low", "medium", "high", "very high".
```{r}
# Define a simple data frame
simple_data_2 <- data.frame( id = 1:10, score = c(55, 67, 22, 87, 45, 36, 91, 75, 59, 66) ) 

# Display the data frame
print(simple_data_2)

# Apply categorize_by_quantiles to categorize "score" into four categories
data_categorized_2 <- categorize_by_quantiles(
    numerical_data = simple_data_2,
    numerical_var_name = "score",
    categorical_labels = c("low", "medium", "high", "very high"),
    num_categories = 4
)

# Display the result
print(data_categorized_2)

```

#### 3. Using the penguins Dataset
This example uses **penguins** dataset and categorize **body_mass_g** into three categories: "light", "average", "heavy".
```{r}
# Categorize body_mass_g into three quantile-based categories
penguins_categorized <- categorize_by_quantiles(
    numerical_data = palmerpenguins::penguins,
    numerical_var_name = "body_mass_g",
    categorical_labels = c("light", "average", "heavy"),
    num_categories = 3
)
# Move the new column body_mass_g_category to the front
penguins_categorized <- dplyr::select(penguins_categorized, body_mass_g_category, dplyr::everything())

# Display the result
head(penguins_categorized)
print(penguins_categorized)
```
